---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r echo=TRUE}
con <- unz ( "activity.zip" , "activity.csv")
activdata<-read.csv( con  )
summary(activdata)
```


## What is mean total number of steps taken per day?
+ Calculate total steps per day  
```{r echo=TRUE}
total_steps_perday<-sapply(split(activdata$steps, activdata$date) ,sum ,na.rm=TRUE)
```

+ Make a histogram of the total number of steps taken each day  
```{r fig.align='left', echo=TRUE}
hist ( total_steps_perday )
```

+ Calculate mean steps taken each day  
```{r echo=TRUE}
mean_steps_perday <- mean (total_steps_perday)
print(mean_steps_perday)
```

+ Calculate median steps taken each day  
```{r echo=TRUE}
med_steps_perday <- median (total_steps_perday)
print (med_steps_perday)
```

## What is the average daily activity pattern?

+ Calculate average steps at each interval
```{r echo=TRUE}
avg_steps_at_interval <- sapply ( split (activdata$steps, activdata$interval), mean, na.rm = TRUE)
```

+ Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r fig.align='left' } 
plot ( names(avg_steps_at_interval), avg_steps_at_interval, type="l", xlab="at interval", ylab="avg steps")
```

+ Find the 5-minute interval, which on average across all the days in the dataset, contains the maximum number of steps:
```{r echo=TRUE}
avg_ints <-  as.data.frame ( cbind (interval = (names(avg_steps_at_interval)) , avg_steps_at_interval = as.data.frame(avg_steps_at_interval) , stringsAsFactors=FALSE ) )
a<-avg_ints$interval[avg_ints$avg_steps_at_interval == max (avg_ints$avg_steps_at_interval)]
print(a)
```

## Imputing missing values
+ Calculate and report the total number of missing values in the dataset
```{r echo=TRUE}
s<- summary( as.factor (activdata$steps ))
na_vals <- s["NA's"]
print ( na_vals )
```

+ Get the mean values for the given interval to replace missing values
```{r echo=TRUE}
avg_ints <-  as.data.frame ( cbind (interval = (names(avg_steps_at_interval)) , avg_steps_at_interval = as.data.frame(avg_steps_at_interval) ) )
head(avg_ints)
```

+ Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r echo=TRUE}
missing_values <- is.na ( activdata$steps)
intervals_w_missing_vals <- activdata$interval[missing_values]
activdata_imp <- activdata
activdata_imp$steps[missing_values] <- avg_ints$avg_steps_at_interval [match ( intervals_w_missing_vals, avg_ints$interval )]
```

Original set:
```{r echo=TRUE}
head(activdata)
```

After imputing:
```{r echo=TRUE}
head(activdata_imp)
```

+ Make a histogram of the total number of steps taken each day using imputed data set
```{r echo=TRUE, fig.align='left'}
total_steps_perday_imp<-sapply(split(activdata_imp$steps, activdata_imp$date) ,sum ,na.rm=TRUE)
hist ( total_steps_perday_imp )
```

+ Calculate and report the mean and median after imputing
```{r echo=TRUE}
mean_steps_perday_imp <- mean (total_steps_perday_imp)
print(mean_steps_perday_imp)
med_steps_perday_imp <- median (total_steps_perday_imp)
print (med_steps_perday_imp)
```

Original mean was:
```{r echo=FALSE}
print (mean_steps_perday)
```

Original median was:
```{r echo=FALSE}
print (med_steps_perday)
```

## Are there differences in activity patterns between weekdays and weekends?
```{r echo=TRUE, fig.align='left'}
activdata_imp <- transform ( activdata_imp, date = as.Date (activdata_imp$date))
w <- ifelse (weekdays(activdata_imp$date) %in% c('Sunday','Saturday') , 'weekend', 'weekday' )
activdata_imp$w <- w
avg_steps_imp_weekend <- sapply ( split (activdata_imp$steps[activdata_imp$w=='weekend'], activdata_imp$interval[activdata_imp$w=='weekend']), mean, na.rm = TRUE)
avg_steps_imp_weekdays <- sapply ( split (activdata_imp$steps[activdata_imp$w=='weekday'], activdata_imp$interval[activdata_imp$w=='weekday']), mean, na.rm = TRUE)

par(mfrow=c(2,1))
plot ( names(avg_steps_at_interval), avg_steps_imp_weekdays, type="l", xlab="at interval", ylab="avg steps", main='Activity at weekdays')
plot ( names(avg_steps_at_interval), avg_steps_imp_weekend, type="l", xlab="at interval", ylab="avg steps", main='Activity at weekends')
```
